Demography and linked selection in wild and domesticated maize

### 2015-08-30 manuscript.tex
some changes to discussion

%JRI suggestions
%no strong evidence for hard sweeps, even considering constraint (GERP).
%although this is not super powerful test as misses sweeps on noncoding and only has power if substantial proportion of missense sites are swept, much previous evidence argues against hard sweeps in maize.  
%popgen studies of local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} show importance of standing variation. 
%even in domestication, few fixed differences between maize and teo \cite{hufford2012}, and of domestication loci that have been well studied from a popgen standpoint only one shows hard sweep \cite{wang2015} while there are multiple examples of standing variation \cite{studer2011, wills2013} or selection on multiple mutations \cite{wills2013}. \jri{maybe others? barren stalk 1 (gallavotti 2004)? ramosa 1 (sigmon 2010)? }
%
%this lack of hard sweeps patterning diversity differs from drosophila \citex and capsella \citex and many species with high Ne \jri{cite corbet-detig}. \jri{add something about corbet-dettig results for maize}
%but is consistent with humans \cite{hernandez2011}
%discussion of whether we expect hard vs soft in maize \cite{messer2013population}.
%$\theta_b = 2N\mu_b$ with $mu_b$ being mutation rate of beneficial alleles of selection coefficient s. 
%if $\theta_b > \frac{1}{log{Ns_b}}$ then adaptation from multiple mutations should be common. 
%if  $\theta_b > \frac{1}{log{2Ns_b}}$ then adaptation from standing neutral variation should be common, or if $\theta > \frac{1}{log(1+\frac{2Ns_b}{2Ns_d + 1}}$ for standing deleterious variation with selection coefficient $s_d$. \jri{i think this latter assumes codominance} codominant?)

%Our findings indicate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize.
%One known example of positive selection on a non-genic mutation involves the \emph{tb1} locus of maize, which is one of the best characterized examples of positive selection on a ``domestication gene'' in any crop \cite{clark2006}. However, the maize \emph{tb1} allele was already present in teosinte before domestication \cite{studer2011}, so this this example is in agreement with our finding that hard sweeps are rare. The \emph{gt1} locus is another well-characterized case of positive selection operating on standing variation at an enhancer region \cite{wills2013}. Instances of selection on standing variation such as these, often called soft sweeps, may be a major contributor to maize patterns of diversity \cite{beissinger2014}, and this could be part of the explanation as to why hard sweeps appear to be so rare, despite obvious morphological differences between maize and teosinte.

@@

Our findings demonstrate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Population genomic studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation as the primary source of adaptive variation. 
Even among well-characterized domestication loci, only a single locus shows evidence of a hard sweep on a missense mutation \cite{wang2015}, but several are consistent with selection on standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}.

The lack of hard sweeps patterning diversity differs from drosophila \cite{sattath2011} and capsella \cite{williamson2014},but is consistent with humans \cite{hernandez2011}. In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this hypothesis into question.}

Another possibility is that maize and teosinte don't show evidence of abundant hard sweeps because instead positive selection tends to operate in the form of soft sweeps. The theory of when soft sweeps are expected was worked out by \cite{messer2013population}. They showed that for $\theta_b = 2N\mu_b$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b > 0$, if $\theta_b > \frac{1}{log({Ns_b})}$, then soft sweeps in the form of adaptation from multiple mutations should be common. Conversely, for neutral sites that somehow become beneficial (e.g. change in environment), if $\theta_b > \frac{1}{log({2Ns_b})}$ then soft sweeps in the form of adaptation from standing neutral variation should be common. Alternatively, for previously deleterious sites that become beneficial, if $\theta > \frac{1}{log \left(1+\frac{2Ns_b}{2Ns_d + 1} \right)}$, where $s_d$ is the previous selection coefficient, soft sweeps again the form of adaptation from standing variation should be common. The implications of this theory are that soft sweeps are common when $\theta_b \geq 1$ \cite{messer2013population}, which can be achieved through a large population size or a high mutation rate. The best estimate of mutation rate for maize is $\mu = 3 \times 10^{-8}$  \cite{clark2005}, which implies that soft sweeps should be common if population size is $\geq$ 16.67 million. \textcolor{red}{According to our demographic analysis, this is larger than that of teosinte but, as discussed below, not out of the question for modern maize.}
\subsection{Demography of domestication}

Although many other authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied  exclusively genic polymorphisms and have made a number of limiting assumptions about the demographic model.  
We show that diversity within genes has been strongly reduced by the effects of linked selection, making genic data inappropriate for estimating demography.
Because levels of diversity alone do not allow independent inference of the strength and duration of a bottleneck \cite{tenaillon2004selection}, previous investigations have instead estimated the ratio of these parameters.  
We have overcome this limitations through the use of the full joint SFS of maize and teosinte and consideration of a more realistic model of exponential growth, enabling us to estimate for the first time the strength of the domestication bottleneck.  
Given that modern maize exhibits $\approx 80\%$ of the diversity of teosinte, our estimate that  the effective population size of the initial maize  population was only $\approx 5\%$ of the ancestral teosinte population seems surprisingly low.
The impact of a strong domestication bottleneck is likely ameliorated to some degree by  maize-teosinte gene flow and the rapid post-domestication growth estimated in our model.

@@

\subsection{Demography influences the efficiency of purifying selection}
The observation that pairwise diversity in maize is less affected by the distance to the nearest gene than in teosinte is reasonable from a long-term evolutionary standpoint. Purifying selection works more efficiently in a large population than in a small one \cite{kimura1984}, so this observation likely reflects that prediction. Since maize bottlenecked before recovering exponentially, the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte.

@@

\jri{need paragraph also talking about how this is important for considering impact of linked selection. cite some recent human papers that are considering both demography and linked selection, but also a number of plant papers that e.g. estimate the DFE from current data or just use the sny/nonsyn graphs and ignore recent demographic history. is capsella a good example? } \textcolor{red}{I agree about this paragraph, but I ran out of time to address. Leave like this and I'll add more after Sept. 11.}

>>>>

%this lack of hard sweeps patterning diversity differs from drosophila \citex and capsella \citex and many species with high Ne \jri{cite corbet-detig}. \jri{add something about corbet-dettig results for maize}
%but is consistent with humans \cite{hernandez2011}
%discussion of whether we expect hard vs soft in maize \cite{messer2013population}.
%$\theta_b = 2N\mu_b$ with $mu_b$ being mutation rate of beneficial alleles of selection coefficient s. 
%if $\theta_b > \frac{1}{log{Ns_b}}$ then adaptation from multiple mutations should be common. 
%if  $\theta_b > \frac{1}{log{2Ns_b}}$ then adaptation from standing neutral variation should be common, or if $\theta > \frac{1}{log(1+\frac{2Ns_b}{2Ns_d + 1}}$ for standing deleterious variation with selection coefficient $s_d$. \jri{i think this latter assumes codominance} codominant?)

@@

Our findings demonstrate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Population genomic studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation as the primary source of adaptive variation. 
Even among well-characterized domestication loci, only a single locus shows evidence of a hard sweep on a missense mutation \cite{wang2015}, but several are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Such soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  
Assuming a mutation rate of $3 \times 10^{-8}$ \citep{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \citep{eyre2007distribution} , this implies that soft sweeps should be prevalent in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \citep[e.g.][]{studer2011}.

\jri{this paragraph needs help} The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011}. 
In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this hypothesis into question.}

\subsection{Demography of domestication}
\jri{section needs help} 
Although many other authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied  exclusively genic polymorphisms and have made a number of limiting assumptions about the demographic model.  
We show that diversity within genes has been strongly reduced by the effects of linked selection, making genic data inappropriate for estimating demography.
Because levels of diversity alone do not allow independent inference of the strength and duration of a bottleneck \cite{tenaillon2004selection}, previous investigations have instead estimated the ratio of these parameters.  
We have overcome this limitations through the use of the full joint SFS of maize and teosinte and consideration of a more realistic model of exponential growth, enabling us to estimate for the first time the strength of the domestication bottleneck.  
Given that modern maize exhibits $\approx 80\%$ of the diversity of teosinte, our estimate that  the effective population size of the initial maize  population was only $\approx 5\%$ of the ancestral teosinte population seems surprisingly low.
The impact of a strong domestication bottleneck is likely ameliorated to some degree by  maize-teosinte gene flow and the rapid post-domestication growth estimated in our model.

@@

\subsection{Demography influences the efficiency of purifying selection}
\jri{section needs help} 
The observation that pairwise diversity in maize is less affected by the distance to the nearest gene than in teosinte is reasonable from a long-term evolutionary standpoint. Purifying selection works more efficiently in a large population than in a small one \cite{kimura1984}, so this observation likely reflects that prediction. Since maize bottlenecked before recovering exponentially, the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte.

@@

\jri{need paragraph also talking about how this is important for considering impact of linked selection. cite some recent human papers that are considering both demography and linked selection, but also a number of plant papers that e.g. estimate the DFE from current data or just use the sny/nonsyn graphs and ignore recent demographic history. is capsella a good example? }
